Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan

### 2020-08-21 analysis/paper/paper.Rmd
minor edits when talking about 14C ages
minor edits about intentional vs mechanical

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. These phases were identified according to chronologically diagnostic artifacts. @Chen2007 reports on 17 radiocarbon ages for Kiwulan, and our Bayesian modeling of these ages shows a consistent result with our artefact-based chronology. However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020]. The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. We also examined excavation depth measurements and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate the three phases. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks. More details for the assignment of different phases are in the Online Supplementary Materials [@WangandMarwick2020].

@@

Whether the shape standardization we found results from craft specialization depends partly on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylistic and functional attributes [@Costin2001; @Costin1995]. Mechanical standardization is related to the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000]. in our case, relative changes in the potential number of producers may be inferred from changes in population size at Kiwulan. According to the Dutch census in 1648 [@Nakamura1938, pp. 12], the population at Kiwulan was large but declined in the Chinese period due to the movement of Indigenous people to the south [@Chen2007]. This change in population corresponds to a decline in ceramic abundance at Kiwulan. Thus, we model CV values as a function of the mass of ceramics using Poisson GLM with a link function. The model suggests that ceramic abundance strongly predicts the CV values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that the more standardized vessel shape of the Chinese period may be influenced by a small population, and thus smaller number of pottery producers, if ceramic abundance can be taken to reflect the population size.

However, intentional standardization due to considerations of function or style could also contribute to the shape standardization in our case. To explore this aspect of the relationship between shape standardization and craft specialization, we investigated the function and surface decoration of the vessels. We used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which were probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. To analyse style, we defined surface impressed decorations, usually consisting of multiple bands of geometric motifs, as types of decorations. If two pots shared the same set of motifs but different arrangements of single bands, we considered them two different types. In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The limited evidence about function, and slight differences in style suggest that intentional standardization may have played a minor role at Kiwulan, but further evidence is required to completely rule out this factor.

Additional insight into craft specialization at Kiwulan comes from the spatial pattern of ceramics, which provides information about potential production units and production areas [@Costin2001]. Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995]. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization. Also, it is more likely that ceramics are household-produced, based on the spatial pattern, and no specific facilities of production are evident [@Chen2007].

@@

This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with significance tests for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Much lower variation in ceramic shape was identified during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behaviors rather than the emergence of discrete groups of producers. Despite a smaller sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops. Instead, the ceramic production is likely to have occurred in households. We speculate that the relatively homogeneous appearance of the vessels may suggest an expression of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, which may be heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in ceramic assemblages.

>>>>

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. These phases were identified according to chronologically diagnostic artifacts. Our Bayesian modeling of eight ages from @Chen2007 shows a consistent result with our artefact-based chronology. However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020]. The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. We also examined excavation depth measurements and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate the three phases. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks. More details for the assignment of different phases are in the Online Supplementary Materials [@WangandMarwick2020].

@@

Whether the shape standardization we found results from craft specialization depends partly on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylistic and functional attributes [@Costin2001; @Costin1995]. Mechanical standardization is related to the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000]. In our case, relative changes in the potential number of producers may be inferred from changes in population size at Kiwulan. According to the Dutch census in 1648 [@Nakamura1938, pp. 12], the population at Kiwulan was large but declined in the Chinese period due to the movement of Indigenous people to the south [@Chen2007]. This change in population corresponds to a decline in ceramic abundance at Kiwulan. Thus, we model CV values as a function of the mass of ceramics using Poisson GLM with a link function. The model suggests that ceramic abundance strongly predicts the CV values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that the more standardized vessel shape of the Chinese period may be influenced by a small population, and thus smaller number of pottery producers, if ceramic abundance can be taken to reflect the population size.

However, intentional standardization due to considerations of function or style could also contribute to the shape standardization in our case. To explore this aspect of the relationship between shape standardization and craft specialization, we investigated the function and surface decoration of the vessels. We used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which were probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. To analyse style, we defined surface impressed decorations, usually consisting of multiple bands of geometric motifs, as types of decorations. If two pots shared the same set of motifs but different arrangements of single bands, we considered them two different types. In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The limited evidence about function, and slight differences in style suggest that intentional standardization may have played only a minor role at Kiwulan, and further evidence is required to completely rule out this factor.

Additional insight into craft specialization at Kiwulan comes from the spatial pattern of ceramics, which provides information about potential production units and production areas [@Costin2001]. Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This is interesting because it contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995]. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of the Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. As population across our three occupation phases declines, we see less clustered distributions of pottery, supporting an interpretation of intentional standardization rather than mechanical standardization. The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007].

@@

This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with significance tests for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Much lower variation in ceramic shape was identified during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization emerges from a combination of mechanical and intentional factors. Discrete groups of producers are not evident, favoring the role of intentional factors in this case. The distribution of ceramics in the Chinese period does not support clustered patterns of manufacturing locations, such as workshops. Instead, ceramic production is likely to have occurred in households. We speculate that the relatively homogeneous appearance of the vessels may suggest an expression of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape. The symbolic value of these shapes may be heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in ceramic assemblages.
